---
editor_options: 
  markdown: 
    wrap: 72
---

## **Linking Diet and Reproductive Success in Northwest Atlantic Grey Seals**

Max Henry-Adams (B00952295) and Peter Braithwaite (B00504978)

Data Analysis STAT 4620/5620 Winter 24-25”

Link to GitHub Repository: <https://github.com/MaxHA17/STAT5620.Project>

### **Abstract:**

**Keywords:** Capital breeding strategy, Generalized Linear Models,
Pinnipeds, Quantitative Fatty AcidSignature Analysis, Reproductive
Ecology

### **Introduction:**

For marine predators, prey distribution may vary unpredictably in time
and space, undergoing seasonal, inter-annual, and longer-term changes in
environmental conditions. Understanding how top predators use different
foraging strategies to navigate and adapt to patchy and unpredictable
prey availability has important implications for individual fitness and
population dynamics, yet studies exploring the consequences of such
decisions on fitness are relatively rare (Nathan et al., 2008).

Vertebrates fuel the costs of reproduction along a continuum of an
income-capital breeding strategy, where animals that use an income
breeding strategy rely on concurrent energy accumulation during the
breeding period as opposed to a capital breeding strategy, where energy
acquired and stored prior to the breeding period is used solely to
finance major reproductive costs (Stephens et al., 2009; Stephens et
al., 2014). A capital breeding strategy is thought to be advantageous
for large animals capable of carrying large energy stores and offers
energetic benefits in areas of patchy or unreliable food availability
and allows parents to direct more resources (e.g. time and energy)
towards offspring, as opposed to foraging, during the breeding and
lactation period (Stephens et al., 2009; Stephens et al.,2014). Given
the importance of pre-breeding and lactation period foraging success and
energy acquisition in capital breeders, these species offer interesting
model systems to study the relationships between diet and reproductive
success.

Phocid seals are wide-ranging, large-bodied, and long-lived marine
predators, with individuals of many species foraging across a wide range
of habitats over large temporal scales (Bowen and Jonsen, 2022). Grey
seals (*Halichoerus grypus*) are a long-lived, sexually dimorphic,
phocid species and are capital breeders. In the Sable Island population,
pregnant females haul out in late December or early January,give birth
to a single pup and nurse that pup for 16-18 days, relying solely on
energy accumulated during the foraging period prior to parturition to
support the costs of lactation (Iverson et al., 1993; Stephens et al.,
2009). At or near the abrupt weaning of that pup, females are mated and
return to sea to replenish spent body reserves and, after a several
month period of delayed implantation, to support gestation and most
importantly preparation for the subsequent December or early January
lactation period. As in all mammals, the costs of lactation, and its
contribution to reproductive success, far exceed the costs of gestation.
Like most other phocids, grey seals are generalist predators (Bowen and
Jonsen, 2022), but females tend to feed on a narrower range of energy
rich species (predominantly sand lance, redfish, and other pelagic
fishes) following the lactation period and expand their prey diversity
in the several months leading up to the December-January breeding and
lactation period (Becket al., 2007). Female grey seals reach sexual
maturity at age 4-5 years and can continue to reproduce as late as into
their early 40s (Bowen et al., 2006). During the brief 16–18-day
lactation period, females utilize approximately 25-38% of parturition
body mass and daily maternal mass loss is a significant predictor of pup
growth rates during lactation, with heavier mothers at parturition
weaning heavier pups (Iverson et al., 1993; Mellish et al., 1999). Pup
survival is positively related to pup weaning mass up to the mean
population weaning mass of 51.5 kg, as predicted by a bigger-is-better
hypothesis for size selective mortality (Bowen et al., 2015). The
phenomena of heavier females weaning larger offspring is evident even
from primiparity where it is believed that heavier females have larger
resource stores that they can mobilize during lactation to produce
larger offspring at weaning (Iverson et al., 1993; Mellish et al., 1999;
Bowen et al., 2006). Following the lactation period, parental care
abruptly ends and the pup is left to fast and survive for several weeks
on the energy acquired during lactation before undertaking its first
foraging trip (Noren et al., 2008). Thus, answering questions relating
to the diet of females prior to parturition will improve our
understanding of the importance of habitat use and energy acquisition as
they relate to life history characteristics, such as reproductive
success, in these capital breeders.

Northwest Atlantic grey seals breeding on Sable Island, Nova Scotia,
tend to be central place foragers and have high fidelity to Sable Island
for breeding and as a haul-out site, making them excellent candidates
for long-term studies (Bowen et al., 2015). Using diet estimations
methods, such as quantitative fatty acid (FA) signature analysis (QFASA;
Iverson et al., 2004, Beck et al., 2007), it becomes possible to explore
the relationship between diet variation and reproductive success at the
individual level. QFASA is used to estimate predator diet by comparing
the FA signatures of predator adipose tissue to those of candidate prey
species (Iverson, 2009; Karnovsky et al., 2012). Predator FA signatures
alone have historically been used to provide a qualitative assessment of
spatial and temporal patterns in diet diversity. However, since many FAs
that are biosynthesized within unique prey species accumulate in
predator adipose tissue over time, evaluating the amount of
prey-specific FAs relative to the predator’s overall FA stores can
provide a quantitative estimate of diet, known as QFASA (Iverson etal.,
2004). For accurate diet estimation, species-specific calibration
coefficients must be experimentally derived to account for the altered
incorporation of prey FAs into the predator adipose tissue due to lipid
metabolic processes within the predator (Iverson et al., 2004; Karnovsky
et al., 2012). When calibration coefficients are derived and the FA
signature of many or all potential prey species in an ecosystem is
known, as in the case of the Northwest Atlantic grey seal range (Iverson
et al., 2004; Beck et al., 2007), QFASA can provide a statistical
approach to accurately depicting diet and prey diversity over
ecologically relevant time scales (e.g., weeks to several months). This
study will test the hypothesis that differences in diet will affect an
individual’s ability to store energy prior to parturition, and thus
influence their reproductive success. Specifically, we will explore two
research questions:

1\. Does variation in diet influence maternal mass change over the
foraging period leading up toparturition in Northwest Atlantic grey
seals?

2\. Does female diet predict pup weaning mass in Northwest Atlantic grey
seals?

### **Data Description**

*Response variables (organized by research question):*

1.  Maternal mass change during the pre-breeding season foraging period:
    continuous variable ranging from -45 – 88 kg. Calculated by
    subtracting the maternal recovery mass from her deployment mass.
    Insert plot of mass change distribution here?

2.  Pup weaning mass: continuous variable ranging from 34.5 – 63 kg.
    Denotes the mass of the pup after the mother has terminated the
    lactation period and departed the pup and/or Sable Island. Insert
    plot of pup weaning mass distribution here?

*Predictor variables*

1.  Year: discrete variable with values ranging from 1996-2015. Denotes
    the year in which each observation was collected.

2.  Maternal ID: discrete variable with a unique identifier for each
    individual female.

3.  Pup sex: categorical; male or female.

4.  Maternal age: discrete variable with values ranging from 7-40 years
    old.

5.  Maternal deployment mass: continuous variable with values ranging
    from 116 – 250 kg. Describes the female mass when she was located
    pregnant on Sable Island 3-6 months prior to parturition and prior
    to undergoing her pre-parturition foraging period.

6.  Maternal mass at parturition: continuous variable ranging from \~
    137 – 250 kg. This is the mass of the female after returning to
    Sable Island and giving birth to her pup.

7.  Maternal dominant prey species: discrete variable with 6 unique
    values. Represents the species making up the greatest proportion of
    the female diet.

8.  Maternal dietary energy density: continuous variable ranging from
    4.88 – 6.51 kJ/g. Energy density (kJ/g) of the female diet is
    calculated from the average energy content per gram of prey body
    tissue.

9.  Maternal diet diversity: continuous variable 0.18 – 0.58, no
    specific units. Diversity is calculated using a standardized
    Shannon-Weaver diversity index to quantify individual diet
    diversity.

### **Methods**

The initial step was data cleaning to ensure all readily apparent data
errors were rectified prior to data analysis and model fitting. Most of
the data errors discovered resulted from the absent of critical data
(i.e. missing covariates) and as a result these observations were
removed from the data-set. As a result, the first question data-set
resulted in 56 observations (n=56) and the second question resulted in
76 observations (n=76). It should also be noted that additional
observations were filtered from the data-set throughout the analysis
process as resulting from outliers with high leverage. A description of
outlier identificantion and removal can be found in the analysis section
below. Thus, the final data-set used for question one had 72 observation
and the final data-set for question two had 50 observations. Once
cleaned and uploaded to RStudio, initial data plots were made using the
‘flexplot’ and ‘plot_explore’ R function we have included in our Git
Hub. Description of 'plot_explore' function created in this R package
can be found within the STAT5620.Project package. These visualizations
facilitated the exploration of variable distributions, specifically the
response variables. From this it was determined that the response
variables for both questions, Maternal Mass Change and Pup Weaning Mass,
were Gaussian distributed. This normality allowed the exploration of the
data through General Linear Model sas well as more complex Generalized
Linear Models and Mixed Models. Although our original analytical
approach proposal Generalized Linear Mixed Models, it felt prudent to
explore the efficiency, interpretability and computational simplify of
Linear Models if the assumption were met and the data suggested
linearity. As a result, both questions were analysed first as Linear
Models (LM), then as Generalized Linear Models (GLM) with Gaussian
distributions, and finally as Generalized Linear Mixed Models (GLMM)
with random effects and fixed effects on some and all the categorical
data. Initially, Maternal ID was to be included as a random effect to
account for repated measured, individual heterogeneity and allow for
population level inferences to be made using these data. However, a lack
of repeated measures (\<10 per question) limited our ability to include
Maternal ID as a random effect and Maternal ID was assumed ot be fixed.
With each model type we began with all the predictor variables prior to
using the backward and forward “Step” function to reduce the number of
variables through elimination based on the p-values for Linear Models
and AIC for GLM and GLMM. Once we had established the best fit model for
both questions using AIC, the performance of the best models were tested
using Cross Validation, question one using 7 folds and question two
using 11 folds.

# **Analysis**

**Research Question 1:** Here we are exploring whether variation in diet
influence maternal mass change over the foraging period leading up to
parturition in Northwest Atlantic grey seals?

The response variable of interest is maternal mass change. We will first
look at the distribution of this variable:

```{r}
#Load the data
here::here("Processed_Data", "Data_Q1.csv")
library(here)
Data_Q1 <- read.csv(here("Processed_Data", "Data_Q1.csv"))
library(readr)
#Data_Q1 <- read_csv("Processed_Data/Data_Q1.csv")
View(Data_Q1)

#Ensure the data types are correctly assigned
Data_Q1$MomID <- as.factor(Data_Q1$MomID)
Data_Q1$Year <- factor(Data_Q1$Year, levels = sort(unique(Data_Q1$Year)), ordered = TRUE)
Data_Q1$Dominant.prey.species <- as.character(Data_Q1$`Dominant.prey.species`)
Data_Q1$`Diet.diversity` <- as.numeric(Data_Q1$`Diet.diversity`)
Data_Q1$`Dietary.energy.density` <- as.numeric(Data_Q1$`Dietary.energy.density`)

summary(Data_Q1)

#rename columns
library(dplyr)
Data_Q1 <- Data_Q1 %>%
  rename(
    DietEngDen = `Dietary.energy.density`,
    DietDiv = `Diet.diversity`,
    DomSpp = `Dominant.prey.species`,
    MassChange = `Mass.change`)

hist(Data_Q1$MassChange)
```

There are a total of n=76 observations and mass change appears to follow
a normal distribution a Gaussian family distribution is a good starting
point for an initial model.

One of the covariates, dominant prey species, is a categorical variable.
We will first check the number of observations in each category:

```{r}
table(Data_Q1$DomSpp)
```

With only one observation of white hake and 3 observations of capelin,
there is likely not enough data to estimate the effect of these species
on the response variable and we are unable to draw meaningful
conclusions. These points also cause extreme residual values. Capelin
and white hake observations will be removed from the data set:

```{r}
Data_Q1 <- Data_Q1 %>%
  filter(DomSpp != "WhiteHake")

Data_Q1 <- Data_Q1 %>%
  filter(DomSpp != "Capelin")
```

The initial model seeks to model maternal mass change as a function of
dominant prey species, diet diversity, dietary energy density, and
deployment year. We will now explore the apparent relationships between
each continuous and categorical covariate using the plot_explore
function built and contained in the STAT5620.Project R package:

```{r}
library(STAT5620.Project)
plot_explore(Data_Q1, response = "MassChange",
                    continuous_vars = c("DietEngDen", "DietDiv"),
                    categorical_vars = c("Year", "DomSpp"))
```

Initially, it appears that diet diversity is negatively related to mass
change, with energy density possibly having a slight negative
relationship. It also appears that animals with cod, redfish or sand
lance as the dominant prey species gain more mass during foraging than
females who primarily prey on Pollock. There does not appear to be any
clear trends in mass change over time.

We will begin with an initial model with all covariates of interest.
Based on the distribution of mass change, we will begin with a simple
model with a gaussian distribution and an identity link function:

```{r}
library(lme4)
mom_mod <- glm(data=Data_Q1, family = gaussian(link="identity"), formula = MassChange ~ DietDiv + DietEngDen + Year + DomSpp)

par(mfrow = c(2, 2))
plot(mom_mod)
```

The residual plots from this initial model show good homoscedasticity
and normality. There is one observation (observation 6) that appears to
be an extreme value. This observation comes from a female that
experiences an 88 kg increase in mass over the pre-breeding period
foraging period, which is well above the mean mass change value of 28.5.
However, none of the covariate values associated with this observation
are abnormal and this point does not have high leverage. As a results,
the decision is to retain this data point since it is not highly
influential and knowledge of the system suggests it is a valid data
point.

We can now use step selection to preform repeated hypothesis tests for
variable selection to determine which covariates best explain maternal
mass change while minimizing model complexity and colinearity between
covariates.

```{r}
step(mom_mod, direction = "both")
```

The lowest AIC model drops the dietary energy density covariate. We will
update the initial model to drop the energy density covariate:

```{r}
mom_mod2 <- glm(formula = MassChange ~ DietDiv + Year + DomSpp, family = gaussian(link = "identity"),
                data = Data_Q1)
plot(mom_mod2)
```

All model residuals still look good. We can proceed to interpreting
model output:

```{r}
summary(mom_mod2)
```

While several years have mass change values that significantly differ
from the expected mass change value, the effect of time is not of direct
interest to our research question and is only present to account for
changes in prey dynamics over time. As a result, the interpretation of
this output is not relevant. However, we see that females who feed
primarily on pollock experience lower mass change than the reference
level, in this case being females who primarily feed on cod. Being an
identity link function, the model is essentially a simple linear model
and there is no transformations applied to the response variable. As
such, coefficient estimates can be interpreted directly. In this case,
individuals who feed on pollock experience 47.5% lower mass change than
individuals who primarily feed on cod. This effect is presented in
figure 1.

```{r}
ggplot(Data_Q1, aes(x = DomSpp, y = MassChange, fill = DomSpp)) +
  geom_boxplot() +
  stat_summary(
    fun = "mean", geom = "errorbar",
    aes(ymax = ..y.., ymin = ..y..),  
    width = 0.75, color = "red", size = 1.2  
  ) + geom_text(x = "Pollock", y = 20, label = "*",  
                aes(x = x, y = y, label = label),  
                color = "black", size = 6, fontface = "bold"
  ) +
  labs(
    title = "Figure 1. Median (black line) and mean (red line) mass change by dominant prey species",
    x = "Prey species",
    y = "Mass change (kg)", fill = "Dominant prey species"
  ) +
  theme_minimal() + theme(
    axis.line = element_line(color = "black", size = 1),  # Add axis lines
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1)
  ) +
  scale_y_continuous(
    breaks = seq(0, max(Data_Q1$MassChange), by = 25) - 50
  )

```

It may also be worth noting that diet diversity is almost significant at
p = 0.0556. While this p-value is not significant, the negative
relationship between diet diversity and mass change may support the
observed effects of dominant prey species, where individuals who
primarily prey on cod, sand lance, or redfish and have a lower diet
diversity due to this focus on a few profitable species are more
successful in gaining mass than individuals who feed on a more diverse
array of prey species which include less profitable species.

We can perform cross-validation to to test this models predictions. In
order to have each fold represent about 10% of the data, we will perform
7 fold cross validation:

```{r}
set.seed(123)

#Number of folds
k <- 7

# Split the data into folds
DatCV <- Data_Q1 %>% mutate(fold = sample(rep(1:k, length.out = n())))

# Create a vector to store RMSE for each fold
rmse_values <- numeric(k)

#Perform cross validation
for (i in 1:k) {
  train_data <- DatCV %>% filter(fold != i)
  test_data  <- DatCV %>% filter(fold == i)
  
  # Fit the model on training data
modCV <- glm(MassChange ~ DietDiv + Year + DomSpp, family = gaussian(link = "identity"), data = Data_Q1)
  
  # Predict on the test data
  predictions <- predict(modCV, newdata = test_data)
  
  # Compute RMSE for this fold
  actuals <- test_data$MassChange
  rmse_values[i] <- sqrt(mean((predictions - actuals)^2))
}

# Summarize
rmse_values
mean_rmse <- mean(rmse_values)
cat("Average RMSE across", k, "folds:", round(mean_rmse, 2), "\n")
```

The average root mean squared error is 15.98, indicating the model
generally predicts values within 16 kg of the true value. Check the
standard deviation to compare the average RMSE to:

```{r}
sd(Data_Q1$MassChange)
```

Compared to the standard deviation of mass change, the RMSE value is
about 40% lower than the standard deviation, meaning the model
predictions are much better than using the population mean to predict
mass change. However, compared to the overall mean mass change within
our dataset (28.5 kg), there is relatively large prediction error. We
can also visualize prediction accuracy with a fitted vs observed value
plot (figure 2):

```{r}
fit_mom <- fitted(mom_mod2)

# Get the observed response variable (actual values)
obs_mom <- Data_Q1$MassChange

# Plot the fitted vs. observed values
ggplot(data = Data_Q1, aes(x = fit_mom, y = obs_mom)) +
  geom_point(color = "black") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +  # Add 1:1 line
  labs(
    title = "Figure 2. Fitted vs. Observed Values with 1:1 line",
    x = "Fitted Values",
    y = "Observed Values"
  ) +
  theme_minimal()

```

While figure 2 shows that the model predictions have consistent accuracy
across the range of fitted values, as suggested by the average RMSE
value obtained from cross-validation, there is reasonably large
prediction error. Given that the model residuals suggest good model fit,
the relatively large prediction error suggest the set of covariates used
in this model are not sufficient in predicting maternal mass change.
From this we can conclude that females who feed on redfish, sand lance,
or cod experience greater mass change than individuals who feed on
Pollock, but other factors outside of maternal diet, such as individual
physiology or characteristics associated with individual foraging
behavior, are responsible for explaining a large proportion of variation
in maternal mass change.

Research question 2:

### Results:

*Question1:*

### **Conclusion:**

### References:

Data source: Sara Iverson laboratory and DFO Grey Seal Program.

Beck, C.A., Iverson, S.J., Bowen, W.D., and Blanchard, W. 2007. Sex
differences in grey seal diet reflect seasonal variation in foraging
behaviour and reproductive expenditure: evidence from quantitative fatty
acid signature analysis. J. Anim. Ecol. 76(3): 490–502.

Bowen, W.D., Iverson, S.J., McMillan, J.I., and Boness, D.J. 2006.
Reproductive performance in grey seals: age-related improvement and
senescence in a capital breeder. J. Anim. Ecol. 75(6): 1340–1351.

Bowen, W.D., den Heyer, C.E., McMillan, J.I., and Iverson, S.J. 2015.
Offspring size at weaning affects survival to recruitment and
reproductive performance of primiparous gray seals. Ecol Evol. 5(7):
1412-1424.

Bowen, W.D. and Jonsen, I.D. 2022. Foraging Ecology and Behavior. In:
Costa, D.P., McHuron, E.A. (eds) Ethology and Behavioral Ecology of
Phocids. Ethology and Behavioral Ecology of Marine Mammals. Springer,
Cham.

Iverson, S.J., Bowen, W.D., Boness, D.J., and Oftedal, O.T. 1993. The
Effect of Maternal Size and Milk Energy Output on Pup Growth in Grey
Seals (Halichoerus grypus). Physiol. Zool. 66(1): 61–88.

Iverson, S.J., Field, C., Bowen, W.D., and Blanchard, W. 2004.
Quantitative fatty acid signature analysis: a new method of estimating
predator diets. Ecol. Monogr. 74(2): 211-235.

Iverson, S.J. 2009. Tracing aquatic food webs using fatty acids: from
qualitative indicators to quantitative determination. In Lipids in
aquatic ecosystems. Edited by M. Arts, M.T. Brett, and M. Kainz.
Springer, New York. pp. 281–308.

Karnovsky, N., Hobson, K.A., and Iverson, S.J. 2012. From lavage to
lipids: Estimating diets of seabirds. Mar. Ecol. Prog. Ser. 451:
263-284.

Mellish, J.E., Iverson, S.J., and Bowen, W.D. 1999. Variation in milk
production and lactation performance in grey seals and consequences for
pup growth and weaning characteristics. Physiol. Biochem. Zool. 72(6):
677–690.

Nathan, R., Getz, W.M., Revilla, E., Holyoak, M., Kadmon, R., Saltz, D.,
and Smouse, P.E. 2008 A movement ecology paradigm for unifying
organismal movement research. Proc. Natl. Acad. Sci. USA. 105(49):
19052–19059.

Noren, S.R., Boness, D.J., Iverson, S.J., McMillan, J.I., and Bowen,
W.D. 2008. Body condition at weaning affects the duration of the
postweaning fast in gray seal pups (Halichoerus grypus). Physiol.
Biochem. Zool. 81: 269–277.

Stephens, P.A., Boyd, I.L., McNamara, J.M. and Houston, A.I. 2009.
Capital breeding and income breeding: their meaning, measurement, and
worth. Ecology. 90(8): 2057-2067. Stephens, P.A., Houston, A.I.,
Harding, K.C., Boyd, I.L. and McNamara, J.M. 2014. Capital and income
breeding: the role of food supply. Ecology. 95(4): 882-896.
